PanGraph: scalable bacterial pan-genome graph construction \\ Supplementary Information

### 2023-02-20 reviewer_response.tex

\reviewer{R1.4}{Throughout the text E. Coli and K. Pneumoniae etc capitalization.}
% \Marco{fixed.}
\response{The capitalization has been corrected.}

\reviewer{R1.5}{Line 213: Core pancontigs have higher than average size - is this what you would expect given that if they are core genes then they will be present in more/all of the samples so there could be more diversity due to being more sampled?}
\Marco{here the pancontig size is the consensus sequence length. I would say that the fact that they have higher-than-average size is mostly due to synteny. Long blocks (several kbps) contain core genes that always appear in the same order, without any accessory sequence ever occurring in these regions. As opposed to accessory regions with a lot of structural diversity, that generates short fragments. But here the main message is that core blocks are well-behaved (close to N50 line in the figure), as opposed to fine-grained structural diversity, part of which contains artifacts generated by the block size limit (many short 100bp blocks). Should we show block frequency distributions in the supplementary?}\\
\Liam{I don't understand their point on diversity - maybe they misunderstood the meaning of size. Maybe pancontig `length' is clearer? I agree with you about synteny and think this is a nice point to add here to explain the observation.}

\reviewer{R1.6}{Figure 6B: could you explore another way to project this data so it is not squished at the ends? I can't read off the fraction agree (shared+private) or disagree which I would like to (although at a glance this is exactly the result you want to see, so happy for it to stand if it is the best projection).}
\Marco{maybe add mean and standard deviation as numbers on the side?}\\
\Liam{that seems ideal.}

\reviewer{R1.7}{Have you tried running PanGraph on more than 300 references? I'd love to see another order of magnitude (not necessary for the paper, just curiosity)}
% \Marco{not really, I never tried on ~500 or ~1000 strains. I could try but would not consider this a priority.}\\
% \Liam{agree that it's not necessary.}
\response{The biggest dataset we tried PanGraph on is a collection of 500 E.coli isolates. We used minimap2 as alignment kernel, asm20 sensitivity and values of $\alpha = 100$, $\beta=20$. This took 18 hours and 12 minutes to complete and had maximum memory usage of 17Gb.}

\reviewer{R1.8}{Discussion: Do you know of/have you tried any mapping approaches which could take this reference as an input. Could you touch on the downstream applications of this graph structure? I would like to see discussion of how the graph can be used downstream - e.g. touching on what software can be used to map/align to such a data structure? Or how it can be used to analysis genome structure.}
\Marco{other than explicitly exporting block sequences and mapping to those, we do not really using the pangenome graph as a reference against which to map new strains... But we could expand the discussion with more ideas on applications to study pangenome organization and evolution. For software for further analysis and exploration I have \href{https://github.com/mmolari/pypangraph}{this package} to explore and interact with graphs in python, but not sure it's worth mentioning.}\\
\Liam{Some ideas that could be mentioned. (1) I think one obvious (bad!) use would be to take the core blocks and map reads to those as a `core reference'. (2) I know pandora has \href{https://github.com/iqbal-lab-org/make_prg/}{\texttt{make\_prg}} which will construct a `pangenome reference graph' from a set of aligned sequences - I didn't try it on pangraph output. (3) For genome structure, exploring `hotspots' in the graph as \href{https://www.nature.com/articles/s41467-017-00808-w}{this paper} does could be done using in/out degree of pancontigs.}

\reviewer{R1.9}{Supplementary Information: simulation - could you comment on how/if homologous recombination is included in this model e.g. within a gene.}
\Marco{homologous recombination is not implemented directly in the model. Only sequence transfer between strains. Genes are not explicitly modeled in the simulation.}\\
\Liam{think just adding a comment in the SI saying this is required.}

\reviewer{R1.10}{Supplementary Information C: You comment that the use of null energy parameters results on average in "more and shorter pancontigs" but also that they result in more diverged sequences being merged into the same pancontig. Why is a more diverse pancontig more likely to become fragmented?}
\Marco{I think that basically the reason is that merging more distant paralogs/orthologs can fragment more the graph. Maybe we could explain this more explicitly in the text.}\\
\Liam{makes sense and explaining this reasoning sounds good.}

### 2023-02-20 supplementary.tex
feat: SI

              \item We check that no prior merges have been performed on the same two pancontigs. Sometimes multiple matches are found in the same pancontig, and in this case we prioritize the ones with higher pseudo-energy. Any following match will be discarded in this step. The same merge might however be performed later on in the self-map phase of the algorithm (see step 4).
              \item Flanking regions of the consensus sequences that are not part of the matches are cut and not merged and stored in separate pancontigs. One exceptions is when one of the flanking regions is shorted than the threshold length $\Lthr$. In this case instead of creating a very short pancontig this region is stored as a small flanking \textbf{insertion (check)} in the merged pancontig.
              \item The remaining homolous region is then a candidate for the merging. However the merging might result in more than a single panconting, depending on the underlying alignment. If the alignment contains insertions or deletions longer thant $\Lthr$, these need to appear as separate matches. To this end the CIGAR string is parsed, and the long match is potentially split in multiple sub-matches, with interposed long insertions or deletions that will be saved as separate pancontigs. \textbf{(what happens with small matches in longer flanking indels?)}
              \item At this point every matching sub-region is processed to produce a single merged pancontig. This is done by expanding the alignments of the two original sub-contigs, and stitching them together following the instructions contained in the CIGAR string of the sub-match. Once the full alignment is constructed the new consensus sequence is calculated using the frequency of nucleotides at each position. Each sequence in the alignment is then encoded as variations over this consensus. This terminates the creation of a new pancontigs. \textbf{(check)}.
              \item The merging is repeated for every matching sub-region of the alignment. Each merged pancontig is shorter in size than the startin pancontigs, since it only represents a sub part of the sequence, but its allignment has depth equal to the sum of the depth of the original pancontigs. The original pancontigs are then removed from the pangenome graph, and no new merges can be performed on them at this step of the algorithm. The newly-created pancontigs are added to the graph and the corresponding paths are updated.

@@

    \item In the final step of the procedure we perform some final grooming of the graph. This includes for example removing transitive edges, i.e. pairs of pancontigs that are always traversed in the same way by all paths. These pairs are merged in a single pancontig. This can happen for example when dealing with circular sequences such as bacterial chromosomes. These sequences are transformed into a single linear block at the beginning of the algorithm, which introduces an artificial cut. When this elementary graph is propagated and the unique block is split, a transitive edge is generated across the artificial cut. We remove it in this step. \textbf{(check.)}

@@

% blocks become progressively deeper but shorter. The energy parameters and the threshold lengt determine when to stop.
% implicit assumption of star-like topology.
% consensus updated but no solving of MSA

>>>>

              \item We check that no prior merges have been performed on the same two pancontigs. Sometimes multiple matches are found on the same pancontig, and in this case we prioritize the ones with higher pseudo-energy. Any following match on the same pancontig will be discarded at this step. It will however be performed later on in the self-map phase of the algorithm (see step 4).
              \item Flanking regions of the consensus sequences that are not part of the matches are cut and not merged and stored in separate pancontigs. One exceptions is when any of these flanking regions is shorter than the threshold length $\Lthr$. In this case instead of creating a very short pancontig the region is stored as an indel in the merged pancontig alignment.
              \item The remaining homolous region is then a candidate for the merging. However the merging might result in more than a single panconting, depending on the underlying alignment. If the alignment contains insertions or deletions longer thant $\Lthr$, these need to appear as separate pancontigs. To this end the CIGAR string is parsed, and the long match is potentially split in multiple sub-matches, with interposed long insertions or deletions that will be saved as separate pancontigs.
              \item At this point every (sub-)match is processed to produce a single merged pancontig. This is done by expanding the alignments of the two original pancontigs, and stitching them together following the instructions contained in the CIGAR string of the match. Once the full alignment is constructed the new consensus sequence is re-evaluated, and each sequence is then encoded as variations over this consensus. This completes the creation of a new pancontig.
              \item The original starting pancontigs are then removed from the pangenome graph, and no new merges can be performed on them at this step of the algorithm. The newly-created pancontigs are added to the graph and the corresponding paths are updated.

@@

    \item In the final step of the procedure we perform some final grooming of the graph. This includes for example removing transitive edges, i.e. pairs of pancontigs that are always traversed in the same way by all paths. These pairs are merged in a single pancontig. This can happen for example when dealing with circular sequences such as bacterial chromosomes. These sequences are transformed into a single linear block at the beginning of the algorithm, which introduces an artificial cut. When this elementary graph is propagated and the unique block is split, a transitive edge is generated across the artificial cut. We remove it in this step.

@@

% pancontigs become progressively deeper but shorter. The energy parameters and the threshold lengt determine when to stop.
% implicit assumption of star-like topology.
% consensus updated but no solving of MSA
